---
title: "Chapter 5: Timbre"
order: 5
---

Cepstrograms of all four songs, structured in per song, per side, total. So first a song and analyses, then of the other from that side, then comparing those. Then the other side and concluding what we see in total.


```{r setup}
#| echo: false
#| warning: false
#| message: false
source("../visualisation_funcs.R")
```

#### Listen to Give me the Loot and Bitch Please
<audio controls style="width: 100%;">
  <source src="../gmtl.m4a" type="audio/mp4">
</audio>

<audio controls style="width: 100%;">
  <source src="../bp.m4a" type="audio/mp4">
</audio>

```{r, fig.width=11, fig.height=6}
#| echo: false
#| warning: false
#| message: false
most_east <- gen_cepstrogram("../data_sv/cepstrogram_gmtl_e.csv", "chebyshev", "Give Me The Loot", "The Notorius B.I.G")
most_west <- gen_cepstrogram("../data_sv/cepstrogram_bitch_please_w_v2.csv", "chebyshev", "Bitch Please", "Snoop Dogg & Xzibit")

grid_plot <- (most_east | most_west)

final_plot_cg <- grid_plot + 
  plot_layout(
    guides = "collect"     # Forces a single legend for the whole patch
  ) + 
  plot_annotation(
    title = "Comparison of Cepstrograms for Most typical E.C. and W.C. track"
  )&
  scale_fill_viridis_c()
  final_plot_cg
```

The cepstrograms reveal distinct differences in timbre complexity between these tracks. "Give Me The Loot" displays a diverse range of MFCC values with multiple peaks across different frequencies, reflecting the sample-rich, layered production characteristic of East Coast hip-hop. "Bitch Please" shows more concentrated spectral energy, particularly in the lower MFCC coefficients, representing a more straightforward, groove-oriented West Coast sound.

#### Listen to C.R.E.A.M and Without Me
<audio controls style="width: 100%;">
  <source src="../cream.m4a" type="audio/mp4">
</audio>

<audio controls style="width: 100%;">
  <source src="../wm.m4a" type="audio/mp4">
</audio>

```{r, fig.width=11, fig.height=6}
#| echo: false
#| warning: false
#| message: false
most_east <- gen_cepstrogram("../data_sv/ceptrogram_cream.csv", "chebyshev", "C.R.E.A.M", "Wu-Tang Clan")
most_west <- gen_cepstrogram("../data_sv/cepstrogram_wm.csv", "chebyshev", "Without Me", "Eminem")

grid_plot <- (most_east | most_west)

final_plot_cg <- grid_plot + 
  plot_layout(
    guides = "collect"     # Forces a single legend for the whole patch
  ) + 
  plot_annotation(
    title = "Comparison of Cepstrograms for Most distinct E.C. and W.C. track"
  )&
  scale_fill_viridis_c()
  final_plot_cg
```

Comparing C.R.E.A.M and Without Me, we observe that both tracks maintain relatively stable timbral characteristics throughout. C.R.E.A.M exhibits consistent MFCC distribution, reflecting Wu-Tang Clan's cohesive sample-based production style. Without Me shows periodic variations in spectral content, corresponding to shifts in the vocal delivery and instrumental background, demonstrating Eminem's more dynamic use of timbre as a production element.
