---
title: "Chapter 7: Classification & Clustering Analysis"
order: 7
---

## Introduction

This chapter presents a comprehensive machine learning analysis of East Coast and West Coast hip-hop tracks using multiple classification and clustering techniques. We begin with unsupervised K-Means clustering to discover natural groupings in the data, followed by supervised K-Nearest Neighbors (KNN) classification to build a predictive model. Finally, we employ both Decision Tree** and Random Forest classifiers to identify the key features that distinguish between regional musical styles. This multi-model approach provides both exploratory insights and actionable feature importance metrics.

---

## Feature Engineering & Selection Methodology

To improve model performance, we refined our feature set through data-driven analysis:

### Feature Selection
We removed noisy features with near-zero correlation to region:
**Removed:** Key (r = -0.006) and Time Signature (r = -0.11) - minimal predictive value
**Retained:** Acousticness (r = 0.40) and Speechiness (r = 0.39) - strongest discriminators
---

```{r setup}
#| echo: false
#| warning: false
#| message: false
library(tidyverse)
library(tidymodels)
library(ggplot2)
library(rpart.plot)
library(cluster)
library(knitr)
library(kableExtra)

source("../k_means_knn.R")
source("../visualisation_funcs.R")

# Load the data once
data_raw <- read_csv("../balanced_hiphop_corpus.csv")
data <- preprocess_hiphop_data(data_raw)

# Color palette from project theme
east_color <- "#1f78b4"  # Blue for East Coast
west_color <- "#ff7f00"  # Orange for West Coast
```

## K-Means Clustering

K-Means clustering is an unsupervised learning technique that partitions the data into k distinct groups based on feature similarity. Unlike supervised methods, it does not use region labels during training—instead, it discovers natural patterns in the data. This allows us to assess how well the algorithm's clusters align with the known geographic regions.

### Clustering Visualization

```{r kmeans-plot}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 8

kmeans_result <- apply_kmeans(data, k = 2)

# Custom visualization with project color scheme
features <- kmeans_result$data %>% select(-where(is.factor), -where(is.character))
pca <- prcomp(features, scale. = TRUE)
pca_data <- as.data.frame(pca$x[, 1:2])
pca_data$region <- kmeans_result$data$region
pca_data$cluster <- kmeans_result$data$cluster

ggplot(pca_data, aes(x = PC1, y = PC2)) +
  stat_ellipse(aes(fill = cluster), geom = "polygon", alpha = 0.15, color = NA) +
  geom_point(aes(color = cluster, shape = region), size = 3.5, alpha = 0.8) +
  scale_color_manual(
    name = "K-Means Cluster",
    values = c("1" = east_color, "2" = west_color),
    labels = c("1" = "Cluster 1", "2" = "Cluster 2")
  ) +
  scale_fill_manual(
    name = "K-Means Cluster",
    values = c("1" = east_color, "2" = west_color),
    labels = c("1" = "Cluster 1", "2" = "Cluster 2")
  ) +
  scale_shape_manual(
    name = "Actual Region",
    values = c("East Coast" = 16, "West Coast" = 17)
  ) +
  theme_minimal(base_size = 12) +
  labs(
    title = "K-Means Clustering: Discovering Natural Patterns",
    subtitle = "Shape indicates actual region | Color indicates assigned cluster",
    x = "Primary Characteristic (PC1)", 
    y = "Secondary Characteristic (PC2)"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```

### Clustering Performance Metrics

```{r kmeans-metrics}
#| echo: false
#| message: false
#| warning: false

# Calculate metrics
silhouette_score <- round(kmeans_result$silhouette, 4)
wcss <- round(sum(kmeans_result$model$withinss), 2)

# Create metrics table
metrics_kmeans <- data.frame(
  Metric = c("Silhouette Score", "Total Within-Cluster Sum of Squares (WCSS)"),
  Value = c(silhouette_score, wcss)
)

kable(metrics_kmeans, 
      col.names = c("Metric", "Value"),
      caption = "K-Means Clustering Performance") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE, position = "left")
```

**Interpretation:** The Silhouette Score ranges from -1 to 1, where higher values indicate better-separated clusters. The WCSS measures the total distance of points from their assigned cluster centers—lower values suggest more compact clusters.

---

## K-Nearest Neighbors (KNN) Classification

K-Nearest Neighbors is a supervised classification algorithm that predicts a song's region by finding the k most similar songs in the training data and using majority voting. Through cross-validation, we identified the optimal value of k that balances bias and variance.

### KNN Predictions

```{r knn-plot}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 8

knn_result <- train_knn(data)

# Get out-of-fold predictions for fair evaluation
oof_pred <- knn_result$oof_predictions %>%
  select(.pred_class)

plot_data_knn <- data
plot_data_knn$Predicted <- oof_pred$.pred_class
plot_data_knn$Actual <- data$region

# PCA for visualization
features_knn <- plot_data_knn %>% 
  select(-where(is.factor), -where(is.character), -Predicted, -Actual)

pca_knn <- prcomp(features_knn, scale. = TRUE)
pca_data_knn <- as.data.frame(pca_knn$x[, 1:2])
pca_data_knn$Predicted <- plot_data_knn$Predicted
pca_data_knn$Actual <- plot_data_knn$Actual

ggplot(pca_data_knn, aes(x = PC1, y = PC2)) +
  stat_ellipse(aes(fill = Predicted), geom = "polygon", alpha = 0.15, color = NA) +
  geom_point(aes(color = Predicted, shape = Actual), size = 3.5, alpha = 0.8) +
  scale_color_manual(
    name = "KNN Prediction",
    values = c("East Coast" = east_color, "West Coast" = west_color)
  ) +
  scale_fill_manual(
    name = "KNN Prediction",
    values = c("East Coast" = east_color, "West Coast" = west_color)
  ) +
  scale_shape_manual(
    name = "Actual Region",
    values = c("East Coast" = 16, "West Coast" = 17)
  ) +
  theme_minimal(base_size = 12) +
  labs(
    title = "KNN Classification: Model Predictions vs. Reality",
    subtitle = "Shape indicates actual region | Color indicates KNN prediction",
    x = "Primary Characteristic (PC1)", 
    y = "Secondary Characteristic (PC2)"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```

### KNN Performance Metrics

```{r knn-metrics}
#| echo: false
#| message: false
#| warning: false

# Calculate performance metrics
confusion_knn <- oof_pred %>%
  bind_cols(Actual = plot_data_knn$Actual) %>%
  rename(Predicted = .pred_class)

tp <- sum(confusion_knn$Predicted == "East Coast" & confusion_knn$Actual == "East Coast")
fp <- sum(confusion_knn$Predicted == "East Coast" & confusion_knn$Actual != "East Coast")
fn <- sum(confusion_knn$Predicted != "East Coast" & confusion_knn$Actual == "East Coast")
tn <- sum(confusion_knn$Predicted != "East Coast" & confusion_knn$Actual != "East Coast")

accuracy_knn <- (tp + tn) / (tp + tn + fp + fn)
precision_knn <- tp / (tp + fp)
recall_knn <- tp / (tp + fn)
best_k <- knn_result$best_k$neighbors

# Create metrics table
metrics_knn <- data.frame(
  Metric = c("Best k value", "Accuracy", "Precision (East Coast)", "Recall (East Coast)"),
  Value = c(best_k, round(accuracy_knn, 4), round(precision_knn, 4), round(recall_knn, 4))
)

kable(metrics_knn, 
      col.names = c("Metric", "Value"),
      caption = "KNN Classification Performance") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE, position = "left")
```

---

## Decision Tree Classification

Decision Trees provide an interpretable classification model by recursively partitioning the feature space based on feature values. This model's main advantage is explainability: we can visualize exactly which features and thresholds the algorithm uses to make predictions, and we can extract feature importance scores.

### Decision Tree Model

```{r tree-plot}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 12
#| fig-height: 8

tree_result <- train_decision_tree(data)

visualize_decision_tree(tree_result$fit, "Decision Tree: Regional Hip-Hop Classification")
```

### Feature Importance

```{r feature-importance}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 6

# Extract the rpart model
rpart_model <- extract_fit_engine(tree_result$fit)

# Get feature importance
importance_df <- data.frame(
  Feature = names(rpart_model$variable.importance),
  Importance = as.numeric(rpart_model$variable.importance)
) %>%
  arrange(desc(Importance)) %>%
  head(10) %>%
  mutate(Feature = factor(Feature, levels = Feature))

ggplot(importance_df, aes(x = reorder(Feature, Importance), y = Importance, fill = Importance)) +
  geom_col(color = "black", linewidth = 0.5) +
  scale_fill_gradient(low = west_color, high = east_color, guide = "none") +
  coord_flip() +
  theme_minimal(base_size = 11) +
  labs(
    title = "Decision Tree Feature Importance",
    subtitle = "Top 10 features distinguishing East Coast from West Coast",
    x = "Feature",
    y = "Importance Score"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 13, hjust = 0.5),
    plot.subtitle = element_text(size = 10, hjust = 0.5),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_line(color = "gray90")
  )
```

### Decision Tree Performance Metrics

```{r tree-metrics}
#| echo: false
#| message: false
#| warning: false

# Get predictions using the final fitted model
tree_pred <- predict(tree_result$fit, new_data = data) %>%
  bind_cols(Actual = data$region) %>%
  rename(Predicted = .pred_class)

# Calculate metrics
tp_tree <- sum(tree_pred$Predicted == "East Coast" & tree_pred$Actual == "East Coast")
fp_tree <- sum(tree_pred$Predicted == "East Coast" & tree_pred$Actual != "East Coast")
fn_tree <- sum(tree_pred$Predicted != "East Coast" & tree_pred$Actual == "East Coast")
tn_tree <- sum(tree_pred$Predicted != "East Coast" & tree_pred$Actual != "East Coast")

accuracy_tree <- (tp_tree + tn_tree) / (tp_tree + tn_tree + fp_tree + fn_tree)
precision_tree <- tp_tree / (tp_tree + fp_tree)
recall_tree <- tp_tree / (tp_tree + fn_tree)

# Best parameters
cost_complexity <- round(tree_result$best_params$cost_complexity, 6)
tree_depth <- tree_result$best_params$tree_depth

# Create metrics table
metrics_tree <- data.frame(
  Metric = c("Cost Complexity Parameter", "Max Tree Depth", "Accuracy", "Precision (East Coast)", "Recall (East Coast)"),
  Value = c(cost_complexity, tree_depth, round(accuracy_tree, 4), round(precision_tree, 4), round(recall_tree, 4))
)

kable(metrics_tree, 
      col.names = c("Metric", "Value"),
      caption = "Decision Tree Classification Performance") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE, position = "left")
```

### Decision Tree Misclassified Songs

```{r tree-misclassified}
#| echo: false
#| message: false
#| warning: false

# Identify misclassified songs
misclassified_tree <- tree_pred %>%
  mutate(row_idx = row_number(), is_correct = Predicted == Actual) %>%
  filter(!is_correct) %>%
  select(row_idx, Actual, Predicted) %>%
  mutate(
    "Song Name" = data_raw$`Track Name`[row_idx],
    "Artist" = data_raw$Artist_Name[row_idx],
    "Actual Region" = Actual,
    "Predicted Region" = Predicted,
    "Acousticness" = round(data_raw$Acousticness[row_idx], 4)
  ) %>%
  select("Song Name", Artist, "Actual Region", "Predicted Region", Acousticness)

if (nrow(misclassified_tree) > 0) {
  kable(misclassified_tree,
        caption = "Songs Misclassified by Decision Tree",
        row.names = FALSE) %>%
    kable_styling(bootstrap_options = "striped", full_width = TRUE)
} else {
  cat("**Perfect Classification:** Decision Tree made no errors!")
}
```

---

## Random Forest Classification

Random Forest is an ensemble learning method that builds multiple decision trees and aggregates their predictions. By combining many trees trained on random subsets of data and features, it reduces overfitting and captures complex non-linear patterns that single trees miss. This is particularly valuable for our hip-hop classification task, where subtle production style interactions matter.

### Random Forest Predictions

```{r rf-plot}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 8

# Check if ranger is installed, if not install it
if (!require("ranger", quietly = TRUE)) {
  install.packages("ranger", quiet = TRUE)
  library(ranger)
}

rf_result <- train_random_forest(data)

# Get out-of-fold predictions
rf_oof_pred <- rf_result$oof_predictions %>%
  select(.pred_class)

plot_data_rf <- data
plot_data_rf$Predicted <- rf_oof_pred$.pred_class
plot_data_rf$Actual <- data$region

# PCA for visualization
features_rf <- plot_data_rf %>% 
  select(-where(is.factor), -where(is.character), -Predicted, -Actual)

pca_rf <- prcomp(features_rf, scale. = TRUE)
pca_data_rf <- as.data.frame(pca_rf$x[, 1:2])
pca_data_rf$Predicted <- plot_data_rf$Predicted
pca_data_rf$Actual <- plot_data_rf$Actual

ggplot(pca_data_rf, aes(x = PC1, y = PC2)) +
  stat_ellipse(aes(fill = Predicted), geom = "polygon", alpha = 0.15, color = NA) +
  geom_point(aes(color = Predicted, shape = Actual), size = 3.5, alpha = 0.8) +
  scale_color_manual(
    name = "Random Forest Prediction",
    values = c("East Coast" = east_color, "West Coast" = west_color)
  ) +
  scale_fill_manual(
    name = "Random Forest Prediction",
    values = c("East Coast" = east_color, "West Coast" = west_color)
  ) +
  scale_shape_manual(
    name = "Actual Region",
    values = c("East Coast" = 16, "West Coast" = 17)
  ) +
  theme_minimal(base_size = 12) +
  labs(
    title = "Random Forest Classification: Ensemble Predictions",
    subtitle = "Shape indicates actual region | Color indicates Random Forest prediction",
    x = "Primary Characteristic (PC1)", 
    y = "Secondary Characteristic (PC2)"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```

### Random Forest Performance Metrics

```{r rf-metrics}
#| echo: false
#| message: false
#| warning: false

# Calculate performance metrics
confusion_rf <- rf_oof_pred %>%
  bind_cols(Actual = plot_data_rf$Actual) %>%
  rename(Predicted = .pred_class)

tp_rf <- sum(confusion_rf$Predicted == "East Coast" & confusion_rf$Actual == "East Coast")
fp_rf <- sum(confusion_rf$Predicted == "East Coast" & confusion_rf$Actual != "East Coast")
fn_rf <- sum(confusion_rf$Predicted != "East Coast" & confusion_rf$Actual == "East Coast")
tn_rf <- sum(confusion_rf$Predicted != "East Coast" & confusion_rf$Actual != "East Coast")

accuracy_rf <- (tp_rf + tn_rf) / (tp_rf + tn_rf + fp_rf + fn_rf)
precision_rf <- tp_rf / (tp_rf + fp_rf)
recall_rf <- tp_rf / (tp_rf + fn_rf)

# Best parameters
mtry_param <- rf_result$best_params$mtry
trees_param <- rf_result$best_params$trees
min_n_param <- rf_result$best_params$min_n

# Create metrics table
metrics_rf <- data.frame(
  Metric = c("Features per split (mtry)", "Number of trees", "Min samples per node", "Accuracy", "Precision (East Coast)", "Recall (East Coast)"),
  Value = c(mtry_param, trees_param, min_n_param, round(accuracy_rf, 4), round(precision_rf, 4), round(recall_rf, 4))
)

kable(metrics_rf, 
      col.names = c("Metric", "Value"),
      caption = "Random Forest Classification Performance") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE, position = "left")
```
