---
title: "Chapter 5: Timbre"
order: 5
---

In this chapter, we take a loot at the timbre features of the four songs.


```{r setup}
#| echo: false
#| warning: false
#| message: false
source("../visualisation_funcs.R")
```

<div class="analyzer-track">
  <span class="track-label">Give Me The Loot - The Notorious B.I.G.</span>
  <audio controls>
    <source src="../gmtl.m4a" type="audio/mp4">
  </audio>
</div>

<div class="analyzer-track">
  <span class="track-label">Bitch Please - Snoop Dogg & Xzibit</span>
  <audio controls>
    <source src="../bp.m4a" type="audio/mp4">
  </audio>
</div>

```{r, fig.width=11, fig.height=6}
#| echo: false
#| warning: false
#| message: false
most_east <- gen_cepstrogram("../data_sv/cepstrogram_gmtl_e.csv", "chebyshev", "Give Me The Loot", "The Notorius B.I.G")
most_west <- gen_cepstrogram("../data_sv/cepstrogram_bitch_please_w_v2.csv", "chebyshev", "Bitch Please", "Snoop Dogg & Xzibit")

grid_plot <- (most_east | most_west)

final_plot_cg <- grid_plot + 
  plot_layout(
    guides = "collect"     # Forces a single legend for the whole patch
  ) + 
  plot_annotation(
    title = "Comparison of Cepstrograms for Most typical E.C. and W.C. track"
  )&
  scale_fill_viridis_c()
  final_plot_cg
```

In this figure, you see the cepstrograms of GMTL and BP. GMTL has a distinctive pattern. mfcc02 is bright yellow, just as mfcc04, while mfcc03 is very dark. This means the audio strongly aligns with the second and fourth coefficients but completely lacks the third. Because these horizontal bands barely fluctuate over time, the visual data tell us that the song relies on a highly consistent, looped sample. However, when listening to the song, it sounds less consistent, due to the many samples fading in and out. In the end of the songs, you hear gunshots. These are visible in the cepstrogram by the vertical stripe at the end.

On the other hand, BP has a more diverse timbre. Although the voices are quite steady, there are samples going in and out often. Interestingly, in the last 50 seconds of the track, the pattern stabilizes and a single instrumental loop becomes highly prominent, which explains why mfcc_03 suddenly becomes much darker and more consistent during that final frame.



<div class="analyzer-track">
  <span class="track-label">C.R.E.A.M. - Wu-Tang Clan</span>
  <audio controls>
    <source src="../cream.m4a" type="audio/mp4">
  </audio>
</div>

<div class="analyzer-track">
  <span class="track-label">Without Me - Eminem</span>
  <audio controls>
    <source src="../wm.m4a" type="audio/mp4">
  </audio>
</div>

```{r, fig.width=11, fig.height=6}
#| echo: false
#| warning: false
#| message: false
most_east <- gen_cepstrogram("../data_sv/ceptrogram_cream.csv", "chebyshev", "C.R.E.A.M", "Wu-Tang Clan")
most_west <- gen_cepstrogram("../data_sv/cepstrogram_wm.csv", "chebyshev", "Without Me", "Eminem")

grid_plot <- (most_east | most_west)

final_plot_cg <- grid_plot + 
  plot_layout(
    guides = "collect"     # Forces a single legend for the whole patch
  ) + 
  plot_annotation(
    title = "Comparison of Cepstrograms for Most distinct E.C. and W.C. track"
  )&
  scale_fill_viridis_c()
  final_plot_cg
```

In this figure, you see the cepstrograms of C.R.E.A.M. and Without Me. C.R.E.A.M. has a distinctive pattern. mfcc_02 is solid bright yellow, and mfcc_04 is also relatively bright, while mfcc_03 is dark. This means the audio strongly aligns with the second and fourth coefficients but largely lacks the third, just like in GMTL. Because these horizontal bands barely fluctuate over time, the visual data confirms that the song relies on a highly consistent, looped sample, giving it that steady, unchanging sonic tone. Which is obviously heard in this songs, in contrast to GMTL.

On the other hand, Without Me has a more diverse timbre. The continuous vertical striping across the lower coefficients shows a constantly shifting texture, meaning there are different instruments and rhythmic elements going in and out often. Interestingly, the rapidly alternating yellow and dark patterns in mfcc_02 and mfcc_03 throughout the entire track highlight a highly dynamic, bouncy production style, which explains why the visual data looks so much more fragmented and active compared to a single, static instrumental loop. This cepstrogram has similarities with BP, in the sense that there is a diverse timbre in both.

### Conclusion
To summarize, both GMTL and CREAM have a steady timbre, which is apperantly distinctive for the E.C. BP and WM have a more diverse timbre, which seems to be distictive for the W.C.