---
title: "Chapter 7: Classification & Clustering Analysis"
order: 7
---

In this chapter, we will look at clustering and multiple classification algoritms. Instead of identifying a coast by hand, it will be interesting to see whether there can be algorithmicly determined to which coast a song belongs. First, we will start with a small feature selection section. Then, we try K-Means clustering. After that, we try KNN, A single descision tree and Random Forrest. 

---

#### Feature Selection
In chapter two we saw that a lot of the distributions of features were the same for both coasts. So, to remove nois for both clustering and classification, we removed `Key`, `Time Signature`, `Energy`, `Liveness`, `Mode`, `Loudness` and `Duration`.


```{r setup}
#| echo: false
#| warning: false
#| message: false
library(tidyverse)
library(tidymodels)
library(ggplot2)
library(rpart.plot)
library(cluster)
library(knitr)
library(kableExtra)

source("../k_means_knn.R")
source("../visualisation_funcs.R")

# Load the data once
data_raw <- read_csv("../balanced_hiphop_corpus.csv")
data <- preprocess_hiphop_data(data_raw)

# Color palette from project theme
east_color <- "#1f78b4"  # Blue for East Coast
west_color <- "#ff7f00"  # Orange for West Coast
```

## K-Means Clustering

K-Means clustering is an unsupervised learning technique that partitions the data into k distinct groups based on feature similarity. As we are trying to find out whether K-Means can distinguish the two coasts, we set the K to 2.

### Clustering Visualization

```{r kmeans-plot}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 8

kmeans_result <- apply_kmeans(data, k = 2)

# Custom visualization with project color scheme
features <- kmeans_result$data %>% select(-where(is.factor), -where(is.character))
pca <- prcomp(features, scale. = TRUE)
pca_data <- as.data.frame(pca$x[, 1:2])
pca_data$region <- kmeans_result$data$region
pca_data$cluster <- kmeans_result$data$cluster

ggplot(pca_data, aes(x = PC1, y = PC2)) +
  stat_ellipse(aes(fill = cluster), geom = "polygon", alpha = 0.15, color = NA) +
  geom_point(aes(color = cluster, shape = region), size = 3.5, alpha = 0.8) +
  scale_color_manual(
    name = "K-Means Cluster",
    values = c("1" = east_color, "2" = west_color),
    labels = c("1" = "Cluster 1", "2" = "Cluster 2")
  ) +
  scale_fill_manual(
    name = "K-Means Cluster",
    values = c("1" = east_color, "2" = west_color),
    labels = c("1" = "Cluster 1", "2" = "Cluster 2")
  ) +
  scale_shape_manual(
    name = "Actual Region",
    values = c("East Coast" = 16, "West Coast" = 17)
  ) +
  theme_minimal(base_size = 12) +
  labs(
    title = "K-Means Clustering East Coast and West Coast",
    subtitle = "Shape indicates actual region,  Color indicates assigned cluster",
    x = "Primary Characteristic (PC1)", 
    y = "Secondary Characteristic (PC2)"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```

This is the with PCA visualised result of K-Means clustering for E.C. and W.C. Before looking at metrics, you can see that the clustering did not do well on disecting east and west. The blue cluster is quite small in relation to the orange cluster and shapes differ throughout. Also in the orange cluster is a lot of differentiation.

#### Clustering Performance Metrics

```{r kmeans-metrics}
#| echo: false
#| message: false
#| warning: false

# Calculate metrics
silhouette_score <- round(kmeans_result$silhouette, 4)
wcss <- round(sum(kmeans_result$model$withinss), 2)

# Create metrics table
metrics_kmeans <- data.frame(
  Metric = c("Silhouette Score"),
  Value = c(silhouette_score)
)

kable(metrics_kmeans, 
      col.names = c("Metric", "Value"),
      caption = "K-Means Clustering Performance") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE, position = "left")
```
The Silhouette Score ranges from -1 to 1, where higher values indicate better-separated clusters. Thus, the silhoutte score achieved indicates weak structure. Or, in other words, there is some structure in the data, but not a lot.

## K-Nearest Neighbors (KNN) Classification

K-Nearest Neighbors is a supervised classification algorithm that predicts a song's region by finding the k most similar songs in the training data and using majority voting. Through 5 fold cross-validation, we identified the optimal value of k that balances bias and variance.

#### KNN Predictions

```{r knn-plot}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 8

knn_result <- train_knn(data)

# Get out-of-fold predictions for fair evaluation
oof_pred <- knn_result$oof_predictions %>%
  select(.pred_class)

plot_data_knn <- data
plot_data_knn$Predicted <- oof_pred$.pred_class
plot_data_knn$Actual <- data$region

# PCA for visualization
features_knn <- plot_data_knn %>% 
  select(-where(is.factor), -where(is.character), -Predicted, -Actual)

pca_knn <- prcomp(features_knn, scale. = TRUE)
pca_data_knn <- as.data.frame(pca_knn$x[, 1:2])
pca_data_knn$Predicted <- plot_data_knn$Predicted
pca_data_knn$Actual <- plot_data_knn$Actual

ggplot(pca_data_knn, aes(x = PC1, y = PC2)) +
  stat_ellipse(aes(fill = Predicted), geom = "polygon", alpha = 0.15, color = NA) +
  geom_point(aes(color = Predicted, shape = Actual), size = 3.5, alpha = 0.8) +
  scale_color_manual(
    name = "KNN Prediction",
    values = c("East Coast" = east_color, "West Coast" = west_color)
  ) +
  scale_fill_manual(
    name = "KNN Prediction",
    values = c("East Coast" = east_color, "West Coast" = west_color)
  ) +
  scale_shape_manual(
    name = "Actual Region",
    values = c("East Coast" = 16, "West Coast" = 17)
  ) +
  theme_minimal(base_size = 12) +
  labs(
    title = "KNN Classification: Model Predictions vs. Reality",
    subtitle = "Shape indicates actual region | Color indicates KNN prediction",
    x = "Primary Characteristic (PC1)", 
    y = "Secondary Characteristic (PC2)"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```



#### KNN Performance Metrics

```{r knn-metrics}
#| echo: false
#| message: false
#| warning: false

library(knitr)
library(kableExtra)

# 1. Combine predictions and actuals
confusion_knn <- oof_pred %>%
  bind_cols(Actual = plot_data_knn$Actual) %>%
  rename(Predicted = .pred_class)

# 2. Generate the Confusion Matrix table
conf_matrix <- table(Actual = confusion_knn$Actual, Predicted = confusion_knn$Predicted)

# Convert to a data frame so kable formats it cleanly as a 2x2 grid
conf_matrix_df <- as.data.frame.matrix(conf_matrix)

# Print the Confusion Matrix
kable(conf_matrix_df, 
      caption = "KNN Confusion Matrix") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  add_header_above(c(" " = 1, "Predicted" = ncol(conf_matrix_df)))

# 3. Calculate performance metrics (Manual calculation)
tp <- sum(confusion_knn$Predicted == "East Coast" & confusion_knn$Actual == "East Coast")
fp <- sum(confusion_knn$Predicted == "East Coast" & confusion_knn$Actual != "East Coast")
fn <- sum(confusion_knn$Predicted != "East Coast" & confusion_knn$Actual == "East Coast")
tn <- sum(confusion_knn$Predicted != "East Coast" & confusion_knn$Actual != "East Coast")

accuracy_knn <- (tp + tn) / (tp + tn + fp + fn)
precision_knn <- tp / (tp + fp)
recall_knn <- tp / (tp + fn)

# 4. Create metrics table (excluding 'K')
metrics_knn <- data.frame(
  Metric = c("Accuracy", "Precision", "Recall"),
  Value = c(round(accuracy_knn, 4), round(precision_knn, 4), round(recall_knn, 4))
)

# Print the Metrics Table
kable(metrics_knn, 
      col.names = c("Metric", "Value"),
      caption = "KNN Classification Performance") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

```
## Decision Tree Classification

Decision Trees provide an interpretable classification model by recursively partitioning the feature space based on feature values. This model's main advantage is explainability: we can visualize exactly which features and thresholds the algorithm uses to make predictions, and we can extract feature importance scores.

### Decision Tree Model

```{r tree-plot}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 12
#| fig-height: 8

tree_result <- train_decision_tree(data)

visualize_decision_tree(tree_result$fit, "Decision Tree: Regional Hip-Hop Classification")
```

### Feature Importance

```{r feature-importance}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 6

# Extract the rpart model
rpart_model <- extract_fit_engine(tree_result$fit)

# Get feature importance
importance_df <- data.frame(
  Feature = names(rpart_model$variable.importance),
  Importance = as.numeric(rpart_model$variable.importance)
) %>%
  arrange(desc(Importance)) %>%
  head(10) %>%
  mutate(Feature = factor(Feature, levels = Feature))

ggplot(importance_df, aes(x = reorder(Feature, Importance), y = Importance, fill = Importance)) +
  geom_col(color = "black", linewidth = 0.5) +
  scale_fill_gradient(low = west_color, high = east_color, guide = "none") +
  coord_flip() +
  theme_minimal(base_size = 11) +
  labs(
    title = "Decision Tree Feature Importance",
    subtitle = "Top 10 features distinguishing East Coast from West Coast",
    x = "Feature",
    y = "Importance Score"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 13, hjust = 0.5),
    plot.subtitle = element_text(size = 10, hjust = 0.5),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_line(color = "gray90")
  )
```

### Decision Tree Performance Metrics

```{r tree-metrics}
#| echo: false
#| message: false
#| warning: false

# Get predictions using the final fitted model
tree_pred <- predict(tree_result$fit, new_data = data) %>%
  bind_cols(Actual = data$region) %>%
  rename(Predicted = .pred_class)

# Calculate metrics
tp_tree <- sum(tree_pred$Predicted == "East Coast" & tree_pred$Actual == "East Coast")
fp_tree <- sum(tree_pred$Predicted == "East Coast" & tree_pred$Actual != "East Coast")
fn_tree <- sum(tree_pred$Predicted != "East Coast" & tree_pred$Actual == "East Coast")
tn_tree <- sum(tree_pred$Predicted != "East Coast" & tree_pred$Actual != "East Coast")

accuracy_tree <- (tp_tree + tn_tree) / (tp_tree + tn_tree + fp_tree + fn_tree)
precision_tree <- tp_tree / (tp_tree + fp_tree)
recall_tree <- tp_tree / (tp_tree + fn_tree)

# Best parameters
cost_complexity <- round(tree_result$best_params$cost_complexity, 6)
tree_depth <- tree_result$best_params$tree_depth

# Create metrics table
metrics_tree <- data.frame(
  Metric = c("Cost Complexity Parameter", "Max Tree Depth", "Accuracy", "Precision (East Coast)", "Recall (East Coast)"),
  Value = c(cost_complexity, tree_depth, round(accuracy_tree, 4), round(precision_tree, 4), round(recall_tree, 4))
)

kable(metrics_tree, 
      col.names = c("Metric", "Value"),
      caption = "Decision Tree Classification Performance") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE, position = "left")
```

### Decision Tree Misclassified Songs

```{r tree-misclassified}
#| echo: false
#| message: false
#| warning: false

# Identify misclassified songs
misclassified_tree <- tree_pred %>%
  mutate(row_idx = row_number(), is_correct = Predicted == Actual) %>%
  filter(!is_correct) %>%
  select(row_idx, Actual, Predicted) %>%
  mutate(
    "Song Name" = data_raw$`Track Name`[row_idx],
    "Artist" = data_raw$Artist_Name[row_idx],
    "Actual Region" = Actual,
    "Predicted Region" = Predicted,
    "Acousticness" = round(data_raw$Acousticness[row_idx], 4)
  ) %>%
  select("Song Name", Artist, "Actual Region", "Predicted Region", Acousticness)

if (nrow(misclassified_tree) > 0) {
  kable(misclassified_tree,
        caption = "Songs Misclassified by Decision Tree",
        row.names = FALSE) %>%
    kable_styling(bootstrap_options = "striped", full_width = TRUE)
} else {
  cat("**Perfect Classification:** Decision Tree made no errors!")
}
```

---

## Random Forest Classification

Random Forest is an ensemble learning method that builds multiple decision trees and aggregates their predictions. By combining many trees trained on random subsets of data and features, it reduces overfitting and captures complex non-linear patterns that single trees miss. This is particularly valuable for our hip-hop classification task, where subtle production style interactions matter.

### Random Forest Predictions

```{r rf-plot}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 8

# Check if ranger is installed, if not install it
if (!require("ranger", quietly = TRUE)) {
  install.packages("ranger", quiet = TRUE)
  library(ranger)
}

rf_result <- train_random_forest(data)

# Get out-of-fold predictions
rf_oof_pred <- rf_result$oof_predictions %>%
  select(.pred_class)

plot_data_rf <- data
plot_data_rf$Predicted <- rf_oof_pred$.pred_class
plot_data_rf$Actual <- data$region

# PCA for visualization
features_rf <- plot_data_rf %>% 
  select(-where(is.factor), -where(is.character), -Predicted, -Actual)

pca_rf <- prcomp(features_rf, scale. = TRUE)
pca_data_rf <- as.data.frame(pca_rf$x[, 1:2])
pca_data_rf$Predicted <- plot_data_rf$Predicted
pca_data_rf$Actual <- plot_data_rf$Actual

ggplot(pca_data_rf, aes(x = PC1, y = PC2)) +
  stat_ellipse(aes(fill = Predicted), geom = "polygon", alpha = 0.15, color = NA) +
  geom_point(aes(color = Predicted, shape = Actual), size = 3.5, alpha = 0.8) +
  scale_color_manual(
    name = "Random Forest Prediction",
    values = c("East Coast" = east_color, "West Coast" = west_color)
  ) +
  scale_fill_manual(
    name = "Random Forest Prediction",
    values = c("East Coast" = east_color, "West Coast" = west_color)
  ) +
  scale_shape_manual(
    name = "Actual Region",
    values = c("East Coast" = 16, "West Coast" = 17)
  ) +
  theme_minimal(base_size = 12) +
  labs(
    title = "Random Forest Classification: Ensemble Predictions",
    subtitle = "Shape indicates actual region | Color indicates Random Forest prediction",
    x = "Primary Characteristic (PC1)", 
    y = "Secondary Characteristic (PC2)"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```

### Random Forest Performance Metrics

```{r rf-metrics}
#| echo: false
#| message: false
#| warning: false

# Calculate performance metrics
confusion_rf <- rf_oof_pred %>%
  bind_cols(Actual = plot_data_rf$Actual) %>%
  rename(Predicted = .pred_class)

tp_rf <- sum(confusion_rf$Predicted == "East Coast" & confusion_rf$Actual == "East Coast")
fp_rf <- sum(confusion_rf$Predicted == "East Coast" & confusion_rf$Actual != "East Coast")
fn_rf <- sum(confusion_rf$Predicted != "East Coast" & confusion_rf$Actual == "East Coast")
tn_rf <- sum(confusion_rf$Predicted != "East Coast" & confusion_rf$Actual != "East Coast")

accuracy_rf <- (tp_rf + tn_rf) / (tp_rf + tn_rf + fp_rf + fn_rf)
precision_rf <- tp_rf / (tp_rf + fp_rf)
recall_rf <- tp_rf / (tp_rf + fn_rf)

# Best parameters
mtry_param <- rf_result$best_params$mtry
trees_param <- rf_result$best_params$trees
min_n_param <- rf_result$best_params$min_n

# Create metrics table
metrics_rf <- data.frame(
  Metric = c("Features per split (mtry)", "Number of trees", "Min samples per node", "Accuracy", "Precision (East Coast)", "Recall (East Coast)"),
  Value = c(mtry_param, trees_param, min_n_param, round(accuracy_rf, 4), round(precision_rf, 4), round(recall_rf, 4))
)

kable(metrics_rf, 
      col.names = c("Metric", "Value"),
      caption = "Random Forest Classification Performance") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE, position = "left")
```
