---
title: "Chapter 7: Clustering"
order: 7
---

Clustering songs unsupervised, and try supervised. Data enhancing, feature importance

```{r setup}
#| echo: false
#| warning: false
#| message: false
library(jsonlite)
library(knitr)
library(kableExtra)

# Load results
results <- fromJSON(file("chapter7_results.json"))
```

## K-Means Clustering Visualization

```{r kmeans-plot}
#| echo: false
#| message: false
#| warning: false
source("../k_means_knn.R")
library(readr)

data <- read_csv("../balanced_hiphop_corpus.csv")
data <- preprocess_hiphop_data(data)
kmeans_result <- apply_kmeans(data, k = 2)
visualize_kmeans(kmeans_result, "K-Means Clustering of East vs West Coast Hip-Hop")
```

**Silhouette Score:** `r round(results$kmeans$silhouette, 3)`  
**Within-Cluster Sum of Squares:** `r round(results$kmeans$wcss, 2)`

## KNN Classification Results
```{r knnresult-plot}
#| echo: false
#| message: false
#| warning: false
# Changed chunk name to knn-plot!
knn_result <- train_knn(data)

# Create the visual
visualize_knn(
  knn_result, 
  original_data = data, 
  title = "KNN Predictions: Model Guesses vs. Reality"
)

```
**Best k:** `r results$knn$best_k$neighbors`

## Decision Tree Visualization

```{r tree-plot}
#| echo: false
#| message: false
#| warning: false
data <- read_csv("../balanced_hiphop_corpus.csv")
data <- preprocess_hiphop_data(data)
tree_result <- train_decision_tree(data)
visualize_decision_tree(tree_result$fit, "Decision Tree for East vs West Coast Hip-Hop Classification")
```

**Best Parameters:** Cost Complexity = `r results$decision_tree$best_params$cost_complexity`, Tree Depth = `r results$decision_tree$best_params$tree_depth`

## XGBoost Model

XGBoost (eXtreme Gradient Boosting) was trained to classify tracks as East Coast or West Coast hip-hop, using Bayesian hyperparameter optimization to find the best configuration.

### Feature Importance

The following table shows the top 10 most important features for the XGBoost model's classification decisions:

```{r xgb-feature-importance}
#| echo: false
#| message: false
#| warning: false

# Get top 10 features
top_features <- results$xgboost$feature_importance %>%
  slice_head(n = 10) %>%
  kable(
    col.names = c("Feature", "Importance"),
    digits = 4,
    caption = "Top 10 Feature Importance Scores"
  ) %>%
  kable_styling(bootstrap_options = "striped")

top_features
```

### Misclassified Songs

Songs that the model incorrectly classified. This can help identify edge cases or patterns the model struggles with:

```{r xgb-misclassified}
#| echo: false
#| message: false
#| warning: false

misclassified <- results$xgboost$misclassified_songs %>%
  kable(
    col.names = c("Track Name", "Artist Name", "Actual Region", "Predicted Region"),
    caption = "Songs Misclassified by XGBoost Model"
  ) %>%
  kable_styling(bootstrap_options = "striped", full_width = TRUE)

misclassified
```