---
title: "Chapter 5: Timbre"
order: 5
---

Cepstrograms of all four songs, structured in per song, per side, total. So first a song and analyses, then of the other from that side, then comparing those. Then the other side and concluding what we see in total.


```{r setup}
#| echo: false
#| warning: false
#| message: false
source("../visualisation_funcs.R")
```

#### Listen to Give me the Loot and Bitch Please
<audio controls style="width: 100%;">
  <source src="../gmtl.m4a" type="audio/mp4">
</audio>

<audio controls style="width: 100%;">
  <source src="../bp.m4a" type="audio/mp4">
</audio>

```{r, fig.width=11, fig.height=6}
#| echo: false
#| warning: false
#| message: false
most_east <- gen_cepstrogram("../data_sv/cepstrogram_gmtl_e.csv", "chebyshev", "Give Me The Loot", "The Notorius B.I.G")
most_west <- gen_cepstrogram("../data_sv/cepstrogram_bitch_please_w_v2.csv", "chebyshev", "Bitch Please", "Snoop Dogg & Xzibit")

grid_plot <- (most_east | most_west)

final_plot_cg <- grid_plot + 
  plot_layout(
    guides = "collect"     # Forces a single legend for the whole patch
  ) + 
  plot_annotation(
    title = "Comparison of Cepstrograms for Most typical E.C. and W.C. track"
  )&
  scale_fill_viridis_c()
  final_plot_cg
```

In this figure, you see the cepstrograms of GMTL and BP. GMTL has a distinctive pattern. mfcc02 is bright yellow, just as mfcc04, while mfcc03 is very dark. This means the audio strongly aligns with the second and fourth coefficients but completely lacks the third. Because these horizontal bands barely fluctuate over time, the visual data confirms that the song relies on a highly consistent, looped sample, giving it that steady, unchanging sonic tone

On the other hand, BP has a more diverse timbre. Although the voices are quite steady, there are samples going in and out often. Interestingly, in the last 50 seconds of the track, the pattern stabilizes and a single instrumental loop becomes highly prominent, which explains why mfcc_03 suddenly becomes much darker and more consistent during that final frame.

#### Listen to C.R.E.A.M and Without Me
<audio controls style="width: 100%;">
  <source src="../cream.m4a" type="audio/mp4">
</audio>

<audio controls style="width: 100%;">
  <source src="../wm.m4a" type="audio/mp4">
</audio>

```{r, fig.width=11, fig.height=6}
#| echo: false
#| warning: false
#| message: false
most_east <- gen_cepstrogram("../data_sv/ceptrogram_cream.csv", "chebyshev", "C.R.E.A.M", "Wu-Tang Clan")
most_west <- gen_cepstrogram("../data_sv/cepstrogram_wm.csv", "chebyshev", "Without Me", "Eminem")

grid_plot <- (most_east | most_west)

final_plot_cg <- grid_plot + 
  plot_layout(
    guides = "collect"     # Forces a single legend for the whole patch
  ) + 
  plot_annotation(
    title = "Comparison of Cepstrograms for Most distinct E.C. and W.C. track"
  )&
  scale_fill_viridis_c()
  final_plot_cg
```

Comparing C.R.E.A.M and Without Me, we observe that both tracks maintain relatively stable timbral characteristics throughout. C.R.E.A.M exhibits consistent MFCC distribution, reflecting Wu-Tang Clan's cohesive sample-based production style. Without Me shows periodic variations in spectral content, corresponding to shifts in the vocal delivery and instrumental background, demonstrating Eminem's more dynamic use of timbre as a production element.
